---
title: "Annotating csRNA-seq TSSs with tssUtils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotating csRNA-seq TSSs with tssUtils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through a complete csRNA-seq annotation run on small
synthetic data: classifying TSSs against transcript and feature
annotations, filtering by expression, defining and quantifying enhancers,
correlating TSS with enhancer expression, and computing per-sample TSS
shape statistics.

```{r}
suppressPackageStartupMessages({
  library(tssUtils)
  library(GenomicRanges)
  library(IRanges)
})
```

## Reference data

In a real run you would load a `TxDb` from disk (for example via
`AnnotationDbi::loadDb()`) and BED files of additional features (TEs,
miRNAs, ATAC peaks, …). For this vignette we build a tiny synthetic
`TxDb` in memory using the `txdbmaker` (or older `GenomicFeatures`) helper.

```{r}
txGR <- GRanges("1",
  IRanges(start = c(1000, 5000, 12000),
    end =      c(2500, 7000, 14000)),
  strand = c("+", "-", "+"),
  type = "mRNA",
  tx_id = 1:3,
  tx_name = c("g1.1", "g2.1", "g3.1"))
TxDb <- if (requireNamespace("txdbmaker", quietly = TRUE)) {
  txdbmaker::makeTxDbFromGRanges(txGR)
} else {
  GenomicFeatures::makeTxDbFromGRanges(txGR)
}
txData <- prepTxData(TxDb)
```

A second feature set, here representing transposable elements:

```{r}
te <- GRanges("1",
  IRanges(start = c(20000, 25000), end = c(20500, 25500)),
  strand = c("+", "-"),
  name = c("TE1", "TE2"))
teData <- prepFeatureData(te, defaultType = "transposable_element")
```

Optional auxiliary peak annotations (any `GRanges` with `mcols$name`):

```{r}
atacPeaks <- GRanges("1",
  IRanges(start = c(900, 24800), end = c(1100, 25200)),
  name = c("atac_1", "atac_2"))
```

## Called TSSs and quantification

The TSSs and CPM matrix you'd normally load from your csRNA-seq pipeline:

```{r}
TSS <- GRanges("1",
  IRanges(start = c(1000, 1500, 5500, 12000, 20000, 25300, 30000),
    width = 30),
  strand = c("+", "-", "-", "+", "+", "-", "+"),
  name = paste0("TSS_", 1:7))

quant <- matrix(c(
   5, 4, 6,
   2, 1, 0,
   3, 4, 5,
   8, 7, 9,
   1, 2, 1,
   6, 5, 7,
   0, 0, 0),
  nrow = length(TSS), byrow = TRUE,
  dimnames = list(TSS$name, c("s1", "s2", "s3")))
```

## Annotate

```{r}
anno <- annotateTSS(TSS, txData,
  featureSets = list(TE = teData),
  featureCoding = c(TE = "teTSS"),
  peakAnnotations = list(ATAC = atacPeaks))
anno[, c("TSS", "TSSTypeCoding", "TSSTypeLocation",
  "TSSFeatOrientation", "FeatureID", "FeatureType")]
```

Filter to expressed TSSs (CPM >= 0.5 in at least 2 samples here, since the
synthetic data has only 3 samples):

```{r}
exprFilt <- filterByExpression(quant, minCpm = 0.5, minSamples = 2)
sum(exprFilt)
```

## Enhancers

```{r}
enhRes <- defineEnhancers(TSS, anno, exprFilt)
enhRes$enh
enhQuant <- quantifyEnhancers(enhRes$enh, TSS, quant)
enhQuant
enhStats <- enhancerStrandStats(enhRes$enh, TSS, quant, exprFilt,
  peakAnnotations = list(ATAC = atacPeaks))
enhStats
```

## TSS-enhancer correlations

```{r}
corrDf <- correlateTSSEnhancers(quant[exprFilt, , drop = FALSE], enhQuant,
  TSS[exprFilt], enhRes$enh, minPCC = 0.5)
corrDf
```

A BEDPE file of the correlation pairs (with the PCC embedded in the name):

```{r}
if (nrow(corrDf)) {
  pairTSS <- TSS[match(corrDf$TSS, TSS$name)]
  pairEnh <- enhRes$enh[match(corrDf$Enhancer, enhRes$enh$name)]
  bedpePath <- tempfile(fileext = ".bedpe")
  writeBEDPE(pairTSS, pairEnh, bedpePath,
    names = paste0("corr", format(round(corrDf$PCC, 2), nsmall = 2),
      "_INT_", seq_len(nrow(corrDf))))
}
```

## Sample-resolved TSS shape

For each sample we'd normally load a stranded bigWig pair via
[`readSignal()`] (or `tssShapeFromBigWig()` for a directory of them).
Here we synthesise a small coverage `GRanges` per sample:

```{r}
makeCov <- function(seed) {
  set.seed(seed)
  pos <- seq(1000, 1029)
  GRanges("1", IRanges(pos, pos), strand = "+",
    score = abs(rnorm(length(pos), mean = 5, sd = 2)))
}
signalList <- list(s1 = makeCov(1), s2 = makeCov(2), s3 = makeCov(3))
shape <- tssShape(TSS[1], signalList,
  percentiles = c(1e-99, 0.1, 0.5, 0.9, 1), minScore = 0)
str(shape, max.level = 1)
```

Aggregate across samples to a single thick-coordinate `GRanges` for BED12
export:

```{r}
shape <- tssShape(TSS, signalList, percentiles = c(0.1, 0.5, 0.9),
  minScore = 0)
thick <- aggregateTSSShape(shape, TSS, method = "median")
thick
```

`tssThickBED()` is a convenience reshaper for hand-built BED12 records;
combine it with `rtracklayer::export.bed()` to write to disk.

```{r}
sessionInfo()
```
