---
title: "Inferring TF-activity changes from TSS shape and motifs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inferring TF-activity changes from TSS shape and motifs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The functions in this vignette lean on a few packages that tssUtils only
# Suggests, so we only evaluate the code when they are all available (which
# means, on a machine without them, you will still see the code but not the
# output).
havePkgs <- all(vapply(c("Biostrings", "Rsamtools", "universalmotif", "glmnet"),
  requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(eval = havePkgs)
```

The shape of a TSS distribution, that is, whether initiation is tightly focused
on one dominant base or spread across a broad region, is set in large part by
which transcription factors (TFs) are acting at the promoter. So a change in
shape between two conditions is a phenotype, and the motifs sitting in the
promoter window give us a set of candidate TF identities to explain it. This
vignette walks through that idea end to end on a small synthetic dataset, using
five functions: `tssShapeMatched()`, `deltaShapeVector()`,
`scanPromoterMotifs()`, `shapeMotifActivity()`, and `motifAnchoredDelta()`.

A word of caution before we start. What follows is a population-level
*associational* inference, not a measurement of occupancy. A motif being present
is not the same as a TF being bound, and shape-based reasoning only sees the
positioning or architectural factors (the ones that move where polymerase
starts), so recruitment-only factors will mostly be invisible here. A nomination
from this analysis is a hypothesis to confirm with an orthogonal signal (an ATAC
footprint, a ChIP, or a motif/TF perturbation), not a conclusion on its own.

```{r}
suppressPackageStartupMessages({
  library(tssUtils)
  library(GenomicRanges)
  library(IRanges)
  library(Biostrings)
  library(Rsamtools)
  library(universalmotif)
})
```

## A small synthetic dataset

In a real run you would bring your own genome (a `BSgenome`, an
`Rsamtools::FaFile`, or an indexed FASTA), a `GRanges` of TSS regions from your
peak caller, and per-sample raw-count 5' tracks read with `readSignal()`. Here,
to keep everything self-contained and fast, we make all of it up.

We build a tiny genome with a promoter every 1000 bp, and plant a TATA box about
28 bp upstream of the first twenty promoters (the other twenty are left as
random sequence, to act as negatives).

```{r}
set.seed(2024)
N <- 40
seqlen <- N * 1000 + 1000
genomeSeq <- paste(sample(c("A", "C", "G", "T"), seqlen, replace = TRUE),
  collapse = "")
tssPos <- (1:N) * 1000

tataProms <- 1:20
for (i in tataProms) substr(genomeSeq, tssPos[i] - 30, tssPos[i] - 25) <- "TATAAA"

fa <- tempfile(fileext = ".fa")
writeXStringSet(DNAStringSet(c("1" = genomeSeq)), fa)
indexFa(fa)
```

The TSS regions are short windows around each promoter. (Note that they carry an
`mcols$name`, which every range-based function in the package uses as a key.)

```{r}
si <- Seqinfo("1", seqlen)
TSS <- GRanges("1", IRanges(tssPos - 20, tssPos + 40), strand = "+",
  name = paste0("prom", seq_len(N)))
seqinfo(TSS) <- si
names(TSS) <- TSS$name
```

Finally the signal. We have two conditions, each with two replicates, carrying
raw integer 5' counts. The control is the same broad pattern everywhere; in the
treatment, the TATA promoters sharpen into a focused peak, while everything else
is left untouched.

```{r}
mkCounts <- function(sharpen) {
  parts <- lapply(seq_len(N), function(i) {
    p <- (tssPos[i] - 20):(tssPos[i] + 40)
    spread <- if (i %in% tataProms && sharpen) 4 else 12
    sc <- as.integer(round(40 * exp(-((p - tssPos[i])^2) / (2 * spread^2))) + 1L)
    g <- GRanges("1", IRanges(p, p), strand = "+")
    g$score <- sc
    g
  })
  out <- sort(do.call(c, parts))
  seqinfo(out) <- si
  out
}
condA <- list(ctrl_rep1 = mkCounts(FALSE), ctrl_rep2 = mkCounts(FALSE))
condB <- list(treat_rep1 = mkCounts(TRUE), treat_rep2 = mkCounts(TRUE))
```

Before going any further it is worth simply looking at the data, since the whole
analysis is really just a way of quantifying what the eye already picks out. The
little helper below pulls the mean per-position sense signal for one promoter
out of a condition, and we plot a TATA promoter next to a non-TATA one.

```{r}
profileOf <- function(signalList, i) {
  p <- (tssPos[i] - 20):(tssPos[i] + 40)
  perSample <- vapply(signalList, function(g) {
    g <- g[strand(g) == "+"]
    v <- numeric(length(p))
    hit <- match(start(g), p)
    ok <- !is.na(hit)
    v[hit[ok]] <- g$score[ok]
    v
  }, numeric(length(p)))
  rowMeans(perSample)
}
```

```{r, fig.width = 7, fig.height = 3.4}
rel <- -20:40
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
for (i in c(1, 21)) {
  a <- profileOf(condA, i)
  b <- profileOf(condB, i)
  plot(rel, a, type = "l", lwd = 2, ylim = c(0, max(a, b)),
    xlab = "Position relative to TSS (bp)", ylab = "Mean 5' counts",
    main = if (i %in% tataProms) "TATA promoter" else "non-TATA promoter")
  lines(rel, b, lwd = 2, col = "firebrick")
  abline(v = 0, col = "grey80")
  legend("topright", bty = "n", lwd = 2, col = c("black", "firebrick"),
    legend = c("control", "treatment"))
}
par(op)
```

(Note that the treatment narrows the TATA promoter into a sharp spike while
leaving the non-TATA promoter essentially alone. That contrast, focusing versus
no change, is the phenotype we are about to put numbers on.)

## Depth-matched shape

The first step is the one it is tempting to skip, and the one that matters most.
Shannon entropy and interquantile width are both biased downward at low tag
counts, so if two conditions differ in coverage, you can manufacture a shape
change out of nothing but sequencing depth. `tssShapeMatched()` guards against
this by downsampling every sample, per promoter, to a common target (by default
the minimum total across samples), recomputing the shape on the subsampled
counts, and averaging over a few draws to settle the sampling noise. It is a
drop-in replacement for `tssShape()`, and it must run on raw counts rather than
RPM-normalized tracks, since you cannot meaningfully subsample a normalized
signal.

We pass `antisense = TRUE` so that the opposite-strand totals come along too,
which we would need if we later wanted to look at directionality. (Note that the
downsampling is random, so we call `set.seed()` first to keep the vignette
reproducible; the same applies to the regression and the meta-profile below.)

```{r}
set.seed(1)
shapeA <- tssShapeMatched(TSS, condA, draws = 5, minCount = 20, antisense = TRUE)
shapeB <- tssShapeMatched(TSS, condB, draws = 5, minCount = 20, antisense = TRUE)
```

## A delta-shape response

`deltaShapeVector()` reduces the two shape objects to a single number per
promoter: the chosen statistic, aggregated across the replicates within each
condition, differenced as B minus A. Here we use the change in Shannon entropy,
where a negative value means initiation became more focused in the treatment.

```{r}
dEntropy <- deltaShapeVector(shapeA, shapeB, metric = "entropy")

c(TATA    = mean(dEntropy[paste0("prom", tataProms)]),
  nonTATA = mean(dEntropy[paste0("prom", setdiff(1:N, tataProms))]))
```

(Note that the TATA promoters carry a clearly negative delta-entropy while the
others sit at zero, which is exactly the signal we planted.) Splitting the
per-promoter values by group makes the separation easy to see:

```{r, fig.width = 5, fig.height = 4}
grp <- ifelse(as.integer(sub("prom", "", names(dEntropy))) %in% tataProms,
  "TATA", "non-TATA")
op <- par(mar = c(4, 4, 2, 1))
stripchart(dEntropy ~ grp, method = "jitter", vertical = TRUE, pch = 19,
  col = c("grey50", "firebrick"), ylab = "delta-entropy (treatment - control)",
  main = "Shape change by promoter group")
abline(h = 0, col = "grey80")
par(op)
```

Other metrics are a single argument away: `"width"` (which needs you to name the
two percentile indices bounding the window, via `pctLo` and `pctHi`),
`"modeShift"`, and `"directionality"` (which needs the `antisense = TRUE` shapes
we built above).

## Scanning the promoter window for motifs

Now we need the candidate identities. `scanPromoterMotifs()` takes a window
around each promoter, defined relative to the dominant TSS, fetches its genomic
sequence, and scans it against a motif library with
`universalmotif::scan_sequences()`. We anchor the window on the dominant
position from `aggregateTSSShape()`, which keeps the scan independent of any one
signal track. For the library we use the real TATA box and a GC-rich decoy, so
we can see whether the analysis tells them apart.

```{r}
motifs <- list(
  create_motif("TATAAA", name = "TATA"),
  create_motif("GGGCGG", name = "GC")
)
dominant <- aggregateTSSShape(shapeA, TSS[rownames(shapeA$shannon)])$thickStart
scan <- scanPromoterMotifs(TSS[rownames(shapeA$shannon)], fa, motifs,
  window = c(-50L, 10L), dominantPos = dominant,
  threshold = 1e-3, thresholdType = "pvalue")

colSums(scan$matrices$presence)
```

The result holds two things: a per-promoter `matrices` list (presence, count,
and best-score) for the regression below, and an `occurrences` `GRanges` (with
each hit's genomic position, strand, offset, and sense/antisense label) for the
meta-profile after that. (Note that a few TATA hits turn up by chance in the
random sequence; that is expected, and is part of what the regularized
regression is built to cope with.)

The `offset` column records where each hit sits relative to the dominant TSS, so
a histogram of it shows the planted TATA box stacking up around its expected
position while the spurious hits scatter:

```{r, fig.width = 5, fig.height = 4}
tataOff <- scan$occurrences$offset[scan$occurrences$motif == "TATA"]
op <- par(mar = c(4, 4, 2, 1))
hist(tataOff, breaks = seq(-50, 10, by = 4), col = "grey80", border = "white",
  xlab = "Motif offset from dominant TSS (bp)", main = "TATA occurrences")
abline(v = -28, col = "firebrick", lwd = 2)
par(op)
```

## Which motif explains the shape change?

`shapeMotifActivity()` regresses the delta-shape response on the motif matrix
with a regularized (elastic-net) model, so that collinear motifs are
deconvolved rather than each grabbing a share of the credit. The signed
coefficient per motif is its estimated shape-activity change, and significance
comes from a permutation test, since glmnet does not hand out usable p-values.

```{r}
set.seed(3)
activity <- shapeMotifActivity(dEntropy, scan$matrices$presence,
  alpha = 0.5, significance = "permutation", nperm = 300)
activity
```

TATA is nominated with a clearly negative coefficient and a small permutation
p-value, while the GC decoy is shrunk to exactly zero. Simply put, the presence
of a TATA box is what tracks with the sharpening we measured, which is the
inference we were after. A signed bar chart of the coefficients makes the
nomination obvious at a glance:

```{r, fig.width = 5, fig.height = 4}
ord <- order(activity$coef)
op <- par(mar = c(4, 5, 2, 1))
cols <- ifelse(activity$pPerm[ord] < 0.05, "firebrick", "grey70")
barplot(activity$coef[ord], names.arg = activity$motif[ord], horiz = TRUE,
  las = 1, col = cols, xlab = "shape-activity coefficient",
  main = "delta-entropy ~ motif presence")
abline(v = 0, col = "grey40")
legend("bottomleft", bty = "n", fill = c("firebrick", "grey70"),
  legend = c("p < 0.05", "n.s."))
par(op)
```

## A position-resolved look at the top motif

A coefficient tells you a motif is associated with the change, but not where
around the motif the change happens. `motifAnchoredDelta()` answers that by
building a fixed window centred and strand-oriented on every occurrence of one
motif, extracting the per-position 5' signal in each condition, and returning
condition B minus condition A aggregated across occurrences. A localized feature
at a stereotyped offset, standing out against a shuffled-window control, is the
signature of a direct local effect.

```{r}
topMotif <- activity$motif[1]
occ <- scan$occurrences[scan$occurrences$motif == topMotif]
set.seed(5)
profile <- motifAnchoredDelta(condA, condB, occ, window = c(60, 60),
  ci = "bootstrap", nboot = 200, control = "shuffle")

head(profile)
```

The return value is a tidy data frame, one row per position, with the aggregated
delta (`deltaMean`) and its confidence band (`deltaLo`, `deltaHi`), the raw
per-condition signal (`signalA`, `signalB`), the antisense delta, and the
control band. It is meant to be plotted directly. On the left we show the two
conditions on their own, and on the right the delta against its confidence band
and the shuffled-window control:

```{r, fig.width = 7.5, fig.height = 3.6, eval = havePkgs}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
with(profile, {
  matplot(distance, cbind(signalA, signalB), type = "l", lty = 1, lwd = 2,
    col = c("grey40", "firebrick"),
    xlab = "Distance from motif (bp)", ylab = "Mean 5' signal",
    main = "Per-condition profiles")
  abline(v = 0, col = "grey80")
  legend("topright", bty = "n", lty = 1, lwd = 2,
    col = c("grey40", "firebrick"), legend = c("control (A)", "treatment (B)"))

  plot(distance, deltaMean, type = "l", lwd = 2, ylim = range(deltaLo, deltaHi),
    xlab = "Distance from motif (bp, motif-5' oriented)",
    ylab = "delta 5' signal (B - A)", main = "Delta and control")
  polygon(c(distance, rev(distance)), c(deltaLo, rev(deltaHi)),
    col = adjustcolor("steelblue", 0.25), border = NA)
  lines(distance, deltaMean, lwd = 2)
  lines(distance, control, col = "firebrick")
  abline(h = 0, v = 0, col = "grey80")
  legend("bottomleft", bty = "n", lty = 1, lwd = c(2, 1),
    col = c("black", "firebrick"), legend = c("delta (B - A)", "shuffle control"))
})
par(op)
```

(Note how the delta sits flat on top of the control everywhere except for the
localized dip a fixed distance from the motif, which is the spatial fingerprint
of the effect.)

If you would rather compare against motif-free promoters than against shifted
windows, `control = "background"` takes a `backgroundGR` of anchor sites and
runs the identical pipeline over them instead.

## In closing

Putting the pieces together, the workflow is: depth-match the shape so a
coverage difference cannot pose as a phenotype, reduce it to a delta-shape
response, scan for the motifs that might explain it, regress to nominate one,
and finally resolve where around that motif the signal actually moved. The
synthetic example recovers the planted biology cleanly, but on real data it is
worth remembering that this only ever nominates a candidate. Motif presence is
not occupancy, the shape lens is blind to recruitment-only factors, and so the
honest next step for any hit is an orthogonal confirmation rather than a
conclusion.

For the upstream steps (calling, classifying, and quantifying TSSs, and the
per-sample `tssShape()` that `tssShapeMatched()` builds on), see the
*Annotating csRNA-seq TSSs with tssUtils* vignette.
```{r, include = FALSE}
unlink(c(fa, paste0(fa, ".fai")))
```
